Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-06-10 paper/main.tex
equivariance + brought up related work + comments

There is an assumption in visualization that the connectivity of the elements in a dataset, which is roughly what topology encodes\cite{wilkinsonGrammarGraphics2005}, are preserved in the visual representation. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitive continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Topological continuity allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that continuity in the visual representation beyond points, lines, areas, and other case by case continuities.

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi and by Troy and M\"{o}ller \cite{chiTaxonomyVisualizationTechniques2000,toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions.

\subsection{Equivariance}
Visualizations that preserve the mathematical structure of the data are called equivariant maps. Equivariant maps are maps where every change to an input has an equivalent change in the output. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011a}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should coresspond to changes in the visualization in a perceptually significant manner. While the data representation introduced in \autoref{sec:atct:io} dispenses with a distinction between containers and the data, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi} and the definition of equivariance introduced in \autoref{sec:artist:equivariance} includes the correspondence between data and visual changes.

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are; where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$. For our purposes, $\dbase$ is a set of points in a mathematical space-such as a metric space, manifold, or euclidean space-that encodes the underlying connectivity of the dataset. For example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. On this set \dbase, we can define subsets of connected elements, called opensets \openset. An openset \openset\ is a generalization of an open interval, meaning an openset contains the points inside a bounded region on X and excludes the points on the boundary. A topology $\mathcal{T}_{\dbase}$ defined on \dbase\ is a collection of opensets such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{TopologyVsTopology,bradleyTopologyCategoricalApproach2020}. The topology is encoding the notion that the connectivity is expected to be preserved regardless of how small an openset or large a union of opensets.

@@

\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
Each section function takes as input a point in the basespace $\dbasepoint$ and returns a single point in the fiber. Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records. Since the fiber space is a topological space, it does not really have restrictions on the dimension of the types it can encode or the structure of those types. For example color can be represented as the cartesian product of three distinct sets, one for each color, or a three dimensional topological space $\mathbb{R} \times \mathbb{R} \times \mathbb{R} = \times \mathbb{R}^{3}$.

@@

A brief introduction to category theory for visualization practitioners is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

>>>>

There is an assumption in visualization that the connectivity of the elements in a dataset are preserved in the visual representation; roughly topology provides a method for encoding this connectivity \cite{wilkinsonGrammarGraphics2005}. Bertin gives the examples of discrete data values should be represented by discrete entities such as points, quantitive continuous data by lines, and surfaces through area marks \cite{bertinSemiologyGraphicsDiagrams2011a}. In \autoref{fig:related-work:continuity:ktypes} the records of the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} weather station table can be considered disconnected from each other and therefore have a 0D continuity. This means each distinct row can be reduced to a discrete point and it is expected  that each rows will be encoded as a discrete visual element, such as a standalone scatter point. Each station can also be considered a time series of weather observation since temperature and pressure are sampled from a 1D continuous measurement space. The connectivity of these observations can be modeled as an interval and the data is encoded as a line since that visual element is also 1D continuous. The measurements at set of stations on any given day is a sparse sampling of a 2D geospatial grid and the connectivity can be modeled as a 2D plane. Instead of interpolation, which is how continuity was preserved in the timeseries plot, in this instance the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved. Topological continuity allows us to encode the connectivity of the data in a uniform manner such that we can then generalize the preservation of that continuity in the visual representation beyond points, lines, areas, and other case by case continuities.

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi and by Troy and M\"{o}ller \cite{chiTaxonomyVisualizationTechniques2000,toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions. The typing system proposed in this paper builds on Butler's proposal of using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

\subsection{Equivariance}
Alongside preserving topology, visualization algorithms are expected to preserve the structure in data values when they translate data to visual representations. The mathematical term for this is equivariance and an equivariant map is a map that transforms input with structure to output with compatible structure. Bertin codified the idea that visual encoding of a data field should match the structure of the field; for example ordered data should be encoded as an ordered visual element or quantitative data as a quantitative visual element \cite{bertinSemiologyGraphicsDiagrams2011a}. This structure mapping is necessary to ensure that the visual representations in graphs are not distorting the data\cite{tufteVisualDisplayQuantitative2001} and graphs are easier to understand when properties match \cite{norman_things_smart}. The expectation of matching structure is generalized as a mapping of a binary operator that acts on the data domain to a binary operator that acts on the graphic domain by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. An invertible binary operator is the basis of Kindlemannn and Scheideggers's algebraic model of visualization\cite{kindlmannAlgebraicProcessVisualization2014}. They propose three specific types of equivariance: that the visualization should not change if the data representation (i.e. the data container) changes, that data maps unambiguously to visual elements \cite{ziemkiewiczEmbeddingInformationVisualization2009}, that changing the data should correspond to changes in the visualization in a perceptually significant manner. In \autoref{sec:atct:sheaves} we introduce a uniform abstract data representation layer, the expectation of unambiguous mappings is expressed in \autoref{sec:atct:xi}, and the definition of equivariance introduced in \autoref{sec:artist:equivariance} includes the correspondence between data and visual changes. In this work, we formally specify the conditions necessary to build structure preserving components using category theory. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Structure Preservation In Software}
\label{sec:related-work:software}
Visualization libraries are in part measured by how expressive the components of the library are, where expressiveness is a measure of which structure preserving mappings a tool can implement \cite{mackinlayAutomaticDesignGraphical1987}. While some visualization tools aim to automate the pairing of data with structure preserving visual representations, such as Tableau\cite{StoltePolaris2002,hanrahanVizQL2006,MackinlayShowme2007}, many visualization libraries leave that choice to the user. Libraries that provide small buildable components \cite{wongsuphasawatNavigatingWideWorld2021}, such as components for making boxes or translating data values to colors, mostly embed their connectivity assumptions in their visual algorithms and allow the user choose almost any component, regardless of structure preservation. Domain specific libraries are designed with the assumption of continuities that are common in the domain \cite{HeerSoftware2006}, and therefore can somewhat restrict the space of choices to algorithms that are appropriate for the data. 
Buildable component libraries such as matplotlib\cite{hunterMatplotlib2DGraphics2007} and D3\cite{bostockDataDrivenDocuments2011} support a wide variety of data types, but at the cost of visual algorithms with  parameters that are tuned to the expected data in a way that leaves the over all library feeling uncohesive. VTK\cite{hanwellVisualizationToolkitVTK2015,geveciVTK2012} provides a language for expressing the topological properties of the data, but does so in a non-uniform manner. On the other hand, the topology of a table as a set of discrete rows, as illustrated in \autoref{fig:related-work:continuity:ktypes}, is assumed by 
A Presentation Tool\cite{mackinlayAutomatingDesignGraphical1986, mackinlayAutomatingDesignGraphical1986} and grammar of graphics\cite{wilkinsonGrammarGraphics2005} and the ggplot\cite{wickhamGgplot2ElegantGraphics2016a}, vega\cite{satyanarayanDeclarativeInteractionDesign2014}, and altair\cite{vanderplasAltairInteractiveStatistical2018} libraries built on these frameworks. Image libraries such as Napari\cite{nicholas_sofroniew_2021_4533308} and ImageJ\cite{schneiderNIHImageImageJ2012} and its humanaties data ImagePlot\cite{studiesCulturevisImageplot2021} plugin assumed that the input is 2D continuous. Networking libraries such as gephi\cite{bastianGephiOpenSource2009} and networkx\cite{HagbergExploringNetwork2008} assume a graph like structure. These domain specific libraries tend to benefit from more cohesive apis, but at the cost of a much more limited set of visualization algorithms than the building block libraries.

@@

The \textcolor{base}{base space} is a topological space \dbasec\ with points $\dbasepoint \in \dbase$ and topology $\mathcal{T}_{\dbase}$. For our purposes, $\dbase$ is a set of points that encodes the underlying connectivity of the dataset. For example, $\dbase$ could be the discrete points, interval, or plane in \autoref{fig:related-work:continuity:ktypes}. On this set \dbase, we can define subsets of connected elements, called opensets \openset. An openset \openset\ is a generalization of an open interval, meaning an openset contains the points inside a bounded region on X and excludes the points on the boundary. A topology $\mathcal{T}_{\dbase}$ defined on \dbase\ is a collection of opensets such that the empty set and \dbase are in $\mathcal{T}_{\dbase}$,and every union and intersection of opensets in $\mathcal{T}_{\dbase}$ are also in $\mathcal{T}_{\dbase}$ \cite{TopologyVsTopology,bradleyTopologyCategoricalApproach2020}. The topology is encoding the notion that the connectivity is expected to be preserved regardless of how small an openset or large a union of opensets.

@@

\begin{equation}
  \label{eq:atct:fb:sections}
  \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} \coloneqq \big\{\dsectionc: \opensetc\rightarrow \dtotalc\restriction_{\opensetc} \; \bigm{\vert} \pi(\dsectionc(\dbasepointc)) = \dbasepointc\;for\, all\; \dbasepointc \in \opensetc \big\} 
\end{equation}
Each section function takes as input a point in the basespace $\dbasepoint$ and returns a single point in the fiber. Extending Spivak's fiber bundle description of a table, we define the values returned by a section as named and typed records. Since the fiber space is a topological space, it does not really have restrictions on the dimension of the types it can encode or the structure of those types. For example, wind has a speed component and a direction component and we can consider those separately $\dfiber_{speed} = \mathbb{R}, \dfiber_{direction} = \mathbb{R}$ or as one fiber$\dfiber_{wind} = \mathbb{R}^{2}$.

@@
